---
title: "Indonesia RASKIN Program (rice for the poor): a PCA Analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

RASKIN (<em>beras miskin</em> or rice for the poor) is one of Indonesia's safety net programs with aims to strengthen food security among poor and vulnerable households. The program has been started since 1998 and still ongoing. It provides a subsidized rice---staple food in Indonesia---and distributed through local officials.

This writing makes a simple analysis using Principal Component Analysis (PCA) to take a look at the RASKIN distribution. This writing is heavily inspired by PCA analysis done by [Julia Silge](https://juliasilge.com/blog/best-hip-hop/).

## Data preparation

Using PCA, I try to understand which households are more likely to ever receive RASKIN. The set of characteristics I used in this analysis are household location (urban or rural), does household only has 5 or less members, house ownership, house characteristics (floor type, wall type, roof type), water source, etc.. Every characteristic is coded as a dummy, with 1 as characteristics that resemble wealthier households and 0 otherwise, such as protected water source as 1 and unprotected water source as 0 (this categorization is subjected to me). 

```{r include=FALSE}
library(tidyverse)

setwd("/Users/rinorsaduaneagusta/Downloads/drive-download-20200517T051151Z-001")

df <- foreign::read.dbf("2014Gab_kor14rt.dbf") %>% 
  tibble()

hhchar <- df %>% 
  transmute(hhid = paste0(B1R1, B1R2, B1R7, B1R8),
            raskin = if_else(B7R1A == 1, 1, 0),
            urban = if_else(B1R5 == 1, 1, 0),
            house_for_living_only = if_else(B6R1 == 1, 1, 0),
            small_members = if_else(B6R2 <= 5, 1, 0),
            own_house = if_else(B6R3 %in% c(1, 2, 6), 1, 0),
            good_roof = if_else(B6R5 %in% c(1:5), 1, 0),
            good_wall = if_else(B6R6 == 1, 1, 0),
            good_floor = if_else(B6R7 %in% c(1:3), 1, 0),
            good_drink_source = if_else(B6R9A %in% c(1:4), 1,
                                  if_else(B6R9A %in% c(5, 6, 8) &
                                            B6R10 == 1, 1, 0)),
            good_cook_water = if_else(B6R11_A %in% c(1:6, 8), 1, 0),
            good_wash_water = if_else(B6R12 %in% c(1:6, 8), 1, 0),
            own_toilet = if_else(B6R13A == 1 & 
                                        B6R13B == 1, 1, 0),
            good_septic_tank = if_else(B6R13C == 1, 1, 0),
            good_light_source = if_else(B6R14A %in% c(1, 2), 1, 0),
            good_cook_energy = if_else(B6R15 %in% c(1:4), 1, 0),
            own_bike = if_else(B7R3A == 1, 1, 0),
            own_motorbike = if_else(B7R3B == 1, 1, 0),
            own_cabletv = if_else(B7R3D == 1, 1, 0),
            own_ac = if_else(B7R3E == 1, 1, 0),
            own_waterheater = if_else(B7R3F == 1, 1, 0),
            own_gastube = if_else(B7R3G == 1, 1, 0),
            own_fridge = if_else(B7R3H == 1, 1, 0),
            own_car = if_else(B7R3J == 1, 1, 0),
            own_cablephone = if_else(B8R1 == 1, 1, 0),
            own_phone = if_else(B8R2A == 1, 1, 0),
            own_desktop = if_else(B8R3A == 1, 1, 0),
            own_laptop = if_else(B8R3B == 1, 1, 0)) %>% 
  distinct(hhid, .keep_all = TRUE) %>%
  drop_na()
```

## Correlation between predictors

```{r message = FALSE}
library(corrr)
```


```{r message = FALSE}
hhchar %>% 
  select(urban:own_laptop) %>% 
  correlate() %>% 
  rearrange() %>% 
  rplot(colours = c("red3", "white", "green4"))
```

Virtually every predictor is positively correlated to each other. This is not surprising as a good aspect of household characteristics will be complemented by another good aspect of household characteristics.

## Principal Component Analysis

To implement PCA, I use recipe from tidymodels. This package is helping me a lot with my workflow (if you ever read this, thanks a lot Ms. Silge!).

```{r message = FALSE}
library(tidymodels)
library(tidytext)
```


```{r}
set.seed(123)

# Put each steps in the recipe
raskin_recipe <- recipe(raskin ~ ., data = hhchar) %>% 
  update_role(hhid, new_role = "id") %>% 
  step_pca(all_predictors())

# Apply each steps in the recipe
raskin_prep <- prep(raskin_recipe)

# Extract results from PCA
raskin_tidied_pca <- tidy(raskin_prep, 1)
```

```{r}
# Extracting standard deviation from the extracted dataframe before
raskin_sdev <- raskin_prep$steps[[1]]$res$sdev

raskin_percent_variation <- raskin_sdev^2 / sum(raskin_sdev^2)

# Cumulative sum of variation plot
tibble(
  component = unique(raskin_tidied_pca$component),
  percent_var = cumsum(raskin_percent_variation)) %>%
  filter(component %in% c("PC1", "PC2", "PC3", "PC4", "PC5")) %>% 
  mutate(component = fct_inorder(component)) %>%
  ggplot(aes(component, percent_var)) +
  geom_col(fill = "green4") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Cumulative Sum of Variation",
       x = "Component", y = "Percentage of variation explained")
```

The bar plot above shows that almost 80 percent variation can be explained by only 5 components. The first component is able to explain more than 75 percent variation. 

```{r}
# Which terms contributed most to each component?
raskin_tidied_pca %>% 
  filter(component %in% c("PC1", "PC2")) %>% 
  mutate(terms = reorder_within(terms, abs(value), component)) %>% 
  ggplot(aes(abs(value), terms, fill = value > 0)) +
  geom_col() +
  facet_wrap(~component, ncol = 1, scales = "free_y") +
  scale_y_reordered() +
  scale_fill_manual(values = c("red3", "green4")) +
  labs(y = NULL, x = "Absolute values of contribution",
       fill = "Positive?")
```

I only show the first and the second component to break down each contributor within each component. Surprisingly, all predictors from the first component--which has the largest proportion in explaining the variance--have negative directions, meaning that households with characteristics resembling a wealthier household are less likely to receive RASKIN. 

```{r}
# The same as above except only the top 5 contributors are showed
raskin_tidied_pca %>% 
  filter(component %in% c("PC1", "PC2")) %>% 
  group_by(component) %>% 
  top_n(5, abs(value)) %>% 
  ungroup() %>% 
  mutate(terms = reorder_within(terms, abs(value), component)) %>% 
  ggplot(aes(abs(value), terms, fill = value > 0)) +
  geom_col() +
  facet_wrap(~component, ncol = 1, scales = "free_y") +
  scale_y_reordered() +
  scale_fill_manual(values = c("red3", "green4")) +
  labs(y = NULL, x = "Values",
       fill = "Positive?")
```

Focusing on the top 5 contributors, we can see that the first component mostly about the size of the household, the physical characteristics of the house, and its utility. While the second component is rather hard to interpret, it is mostly about whether the households have a fridge or not, and has a negative direction. So, the conclusion is pretty much the same as the first component, although the terms whether the house is only used for living seems contradicted (but, remember that the variation explained by the second component is less than 5 percent).

How are the households distributed in the plane of the first components? 

```{r}
# Two dimensions plot of the second and the first component
juice(raskin_prep) %>% 
  mutate(raskin = if_else(raskin == 1, "Yes", "No")) %>% 
  ggplot(aes(PC1, PC2, color = as.factor(raskin))) +
  geom_jitter(alpha = 0.2) +
  scale_color_manual(values = c("red3", "green4")) +
  labs(color = "Receive RASKIN?")
```

Let us focus on the direction of the first component. Wealthier characteristics are to the left and poorer characteristics are to the right. We can see that households that have ever received RASKIN tend to be at the right side of the x-axis, although some of the wealthier households also enjoy RASKIN. These results show suggestive evidence that RASKIN is indeed distributed to poor households, but it is far from perfect. While the selection process of RASKIN needs to be improved, the fact that some poorer households never received RASKIN means that the information of this program also has to be promoted more widely and more inclusive. 

This analysis is without limitations. Please note that this is my first time using PCA. My intention is to apply what I've learned so far about this tool, so don't take anything here for granted. Suggestions, critiques, and comments are welcomed!




